---
title: "Week 7 modules"
output: html_notebook
---

#Module 11 - Type I and Type II Error and Power


What is True	  What We Decide	  Result
H0	          H0	          Correctly ‘accept’ the null
H0H0	          HAHA           	Falsely reject the null (Type I error)
HAHA	          H0H0	          Falsely ‘accept’ the null (Type II error)
HAHA          	HAHA          	Correctly reject the null

##Type I Error

```{r}
typeI <- function(mu0, sigma, n, alternative = "two.tailed", alpha = 0.05, k = 10000) {
}
```

```{r}
typeI <- function(mu0, sigma, n, alternative = "two.tailed", alpha = 0.05, k = 1000) {
    p <- rep(NA, k)  # sets up a vector of empty p values
    for (i in 1:k) {
        # sets up a loop to run k simulations
        x <- rnorm(n = n, mean = mu0, sd = sigma)  # draws a sample from our distribution
        m <- mean(x)  # calculates the mean
        s <- sd(x)  # calculates the standard deviation
        z <- (m - mu0)/(s/sqrt(n))  # calculates the T statistic for the sample drawn from the null distribution relative to the null distribution
        # alternatively use t <- (m-mu0)/(s/sqrt(n))
        if (alternative == "less") {
            p[i] <- pnorm(z, lower.tail = TRUE)  # calculates the associated p value
            # alternatively, use p[i] <- pt(t,df=n-1,lower.tail=TRUE)
        }
        if (alternative == "greater") {
            p[i] <- pnorm(z, lower.tail = FALSE)  # calculates the associated p value
            # alternatively, use p[i] <- pt(t,df=n-1,lower.tail=FALSE)
        }
        if (alternative == "two.tailed") {
            if (z > 0) 
                {
                  p[i] <- 2 * pnorm(z, lower.tail = FALSE)
                }  # alternatively, use if (t > 0) {p[i] <- pt(t,df=n-1,lower.tail=FALSE)}
            if (z < 0) 
                {
                  p[i] <- 2 * pnorm(z, lower.tail = TRUE)
                }  # alternatively, use if (t < 0) {p[i] <- pt(t,df=n-1,lower.tail=TRUE)}
        }
    }
    
    curve(dnorm(x, mu0, sigma/sqrt(n)), mu0 - 4 * sigma/sqrt(n), mu0 + 4 * sigma/sqrt(n), 
        main = paste("Sampling Distribution Under the Null Hypothesis\nType I error rate from simulation = ", 
            length(p[p < alpha])/k, sep = ""), xlab = "x", ylab = "Pr(x)", col = "red", 
        xlim = c(mu0 - 4 * sigma/sqrt(n), mu0 + 4 * sigma/sqrt(n)), ylim = c(0, 
            dnorm(mu0, mu0, sigma/sqrt(n))))
    abline(h = 0)
    
    if (alternative == "less") {
        polygon(cbind(c(mu0 - 4 * sigma/sqrt(n), seq(from = mu0 - 4 * sigma/sqrt(n), 
            to = mu0 - qnorm(1 - alpha) * sigma/sqrt(n), length.out = 100), 
            mu0 - qnorm(1 - alpha) * sigma/sqrt(n))), c(0, dnorm(seq(from = mu0 - 
            4 * sigma/sqrt(n), to = mu0 - qnorm(1 - alpha) * sigma/sqrt(n), 
            length.out = 100), mean = mu0, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        q <- pnorm(mu0 - qnorm(1 - alpha) * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n)) - 
            pnorm(mu0 - 4 * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n))
    }
    if (alternative == "greater") {
        polygon(cbind(c(mu0 + qnorm(1 - alpha) * sigma/sqrt(n), seq(from = mu0 + 
            qnorm(1 - alpha) * sigma/sqrt(n), to = mu0 + 4 * sigma/sqrt(n), 
            length.out = 100), mu0 + 4 * sigma/sqrt(n))), c(0, dnorm(seq(from = mu0 + 
            qnorm(1 - alpha) * sigma/sqrt(n), to = mu0 + 4 * sigma/sqrt(n), 
            length.out = 100), mean = mu0, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        q <- pnorm(mu0 + 4 * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n)) - 
            pnorm(mu0 + qnorm(1 - alpha) * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n))
    }
    if (alternative == "two.tailed") {
        polygon(cbind(c(mu0 - 4 * sigma/sqrt(n), seq(from = mu0 - 4 * sigma/sqrt(n), 
            to = mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n), length.out = 100), 
            mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n))), c(0, dnorm(seq(from = mu0 - 
            4 * sigma/sqrt(n), to = mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n), 
            length.out = 100), mean = mu0, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        polygon(cbind(c(mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n), seq(from = mu0 + 
            qnorm(1 - alpha/2) * sigma/sqrt(n), to = mu0 + 4 * sigma/sqrt(n), 
            length.out = 100), mu0 + 4 * sigma/sqrt(n))), c(0, dnorm(seq(from = mu0 + 
            qnorm(1 - alpha/2) * sigma/sqrt(n), to = mu0 + 4 * sigma/sqrt(n), 
            length.out = 100), mean = mu0, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        q <- pnorm(mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n)) - 
            pnorm(mu0 - 4 * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n)) + 
            pnorm(mu0 + 4 * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n)) - 
            pnorm(mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n), mean = mu0, sd = sigma/sqrt(n))
    }
    # print(round(q,digits=3)) # this prints area in the shaded portion(s) of
    # the curve
    return(length(p[p < alpha])/k)  # returns the proportion of simulations where p < alpha
}
```
```{r}
eI <- typeI(mu0 = -3, sigma = 2, n = 5000, alternative = "greater", alpha = 0.05)
```

```{r}
eI <- typeI(mu0 = 5, sigma = 2, n = 1000, alternative = "less", alpha = 0.01)
```
 Bonferroni correction
```{r}
alpha <- 0.05
pvals <- c(1e-04, 0.003, 0.005, 0.01, 0.02, 0.04, 0.045, 0.11, 0.18, 0.23)
sig <- pvals <= alpha/length(pvals)
sig  # first 3 values are less than the adjusted alpha
```


Benjamini & Hochberg correction
```{r}
library(ggplot2)
alpha <- 0.05
psig <- NULL
pvals <- c(1e-04, 0.003, 0.005, 0.01, 0.02, 0.04, 0.045, 0.11, 0.18, 0.27)
for (i in 1:length(pvals)) {
    psig[i] <- alpha * i/length(pvals)
}
d <- data.frame(cbind(rank = c(1:10), pvals, psig))
p <- ggplot(data = d, aes(x = rank, y = pvals)) + geom_point() + geom_line(aes(x = rank, 
    y = psig))
p

```

```{r}
sig <- pvals <= psig  # vector of significant pvalues
sig  # first 5 values are less than the adjusted alpha
```

```{r}
sig <- p.adjust(pvals, method = "bonferroni") <= 0.05
sig  # first 3 adjusted p values are less alpha
```

```{r}
sig <- p.adjust(pvals, method = "BH") <= 0.05
sig  # first 4 adjusted p values are less alpha
```
##Type II Error
```{r}
typeII <- function(mu0, muA, sigma, n, alternative = "two.tailed", alpha = 0.05, 
    k = 1000) {
    p <- rep(NA, k)  # sets up a vector of empty p values
    for (i in 1:k) {
        x <- rnorm(n = n, mean = muA, sd = sigma)  # draw from Ha
        m <- mean(x)
        s <- sd(x)
        z <- (m - mu0)/(s/sqrt(n))  # calculates the Z statistic for the sample drawn from Ha relative to the null distribution
        if (alternative == "less") {
            p[i] <- pnorm(z, lower.tail = TRUE)  # calculates the associated p value
            hyp <- "muA < mu0"
        }
        if (alternative == "greater") {
            p[i] <- pnorm(z, lower.tail = FALSE)
            hyp <- "muA > mu0"
        }
        if (alternative == "two.tailed") {
            if (z > 0) {
                p[i] <- 2 * pnorm(z, lower.tail = FALSE)
            }
            if (z < 0) {
                p[i] <- 2 * pnorm(z, lower.tail = TRUE)
            }
            hyp <- "muA ≠ mu0"
        }
    }
    
    curve(dnorm(x, mu0, sigma/sqrt(n)), mu0 - 4 * sigma/sqrt(n), mu0 + 4 * sigma/sqrt(n), 
        main = paste("Sampling Distributions Under the Null (red)\nand Alternative Hypotheses (blue)\nType II error rate from simulation = ", 
            length(p[p >= alpha])/k, sep = ""), xlab = "x", ylab = "Pr(x)", 
        col = "red", xlim = c(min(c(mu0 - 4 * sigma/sqrt(n), muA - 4 * sigma/sqrt(n))), 
            max(c(mu0 + 4 * sigma/sqrt(n), muA + 4 * sigma/sqrt(n)))), ylim = c(0, 
            max(c(dnorm(mu0, mu0, sigma/sqrt(n))), dnorm(muA, muA, sigma/sqrt(n)))))
    
    curve(dnorm(x, muA, sigma/sqrt(n)), muA - 4 * sigma/sqrt(n), muA + 4 * sigma/sqrt(n), 
        col = "blue", add = TRUE)
    abline(h = 0)
    
    if (alternative == "less") {
        polygon(cbind(c(mu0 - qnorm(1 - alpha) * sigma/sqrt(n), seq(from = mu0 - 
            qnorm(1 - alpha) * sigma/sqrt(n), to = muA + 4 * sigma/sqrt(n), 
            length.out = 100), muA + 4 * sigma/sqrt(n))), c(0, dnorm(seq(mu0 - 
            qnorm(1 - alpha) * sigma/sqrt(n), to = muA + 4 * sigma/sqrt(n), 
            length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        abline(v = mu0 - qnorm(1 - alpha) * sigma/sqrt(n), col = "black", lty = 3, 
            lwd = 2)
    }
    
    if (alternative == "greater") {
        polygon(cbind(c(muA - 4 * sigma/sqrt(n), seq(from = muA - 4 * sigma/sqrt(n), 
            to = mu0 + qnorm(1 - alpha) * sigma/sqrt(n), length.out = 100), 
            mu0 + qnorm(1 - alpha) * sigma/sqrt(n))), c(0, dnorm(seq(from = muA - 
            4 * sigma/sqrt(n), to = mu0 + qnorm(1 - alpha) * sigma/sqrt(n), 
            length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0), border = "black", 
            col = "grey")
        abline(v = mu0 + qnorm(1 - alpha) * sigma/sqrt(n), col = "black", lty = 3, 
            lwd = 2)
    }
    
    if (alternative == "two.tailed") {
        abline(v = mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n), col = "black", 
            lty = 3, lwd = 2)
        abline(v = mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n), col = "black", 
            lty = 3, lwd = 2)
        
        if (z > 0) {
            # greater
            polygon(cbind(c(muA - 4 * sigma/sqrt(n), seq(from = muA - 4 * sigma/sqrt(n), 
                to = mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n), length.out = 100), 
                mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n))), c(0, dnorm(seq(from = muA - 
                4 * sigma/sqrt(n), to = mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n), 
                length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0), border = "black", 
                col = "grey")
        }
        
        # less
        if (z < 0) {
            polygon(cbind(c(mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n), seq(from = mu0 - 
                qnorm(1 - alpha/2) * sigma/sqrt(n), to = muA + 4 * sigma/sqrt(n), 
                length.out = 100), muA + 4 * sigma/sqrt(n))), c(0, dnorm(seq(mu0 - 
                qnorm(1 - alpha/2) * sigma/sqrt(n), to = muA + 4 * sigma/sqrt(n), 
                length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0), border = "black", 
                col = "grey")
        }
    }
    
    return(length(p[p >= alpha])/k)
}
```

```{r}
eII <- typeII(mu0 = 2, muA = 4, sigma = 3, n = 6, alternative = "greater")  # Ha > H0
```
Questions: is n (Type II error function), refering to number in each sample, or all together
```{r}
eII <- typeII(mu0 = 5, muA = 2, sigma = 4, n = 18, alternative = "less")  # Ha < H0
```
##Power
Power is the probability of correctly rejecting a null hypothesis that is untrue. For a test that has a Type II error rate of ββ, the statistical power is defined, simply, as 1−β1−β. Power values of 0.8 or greater are conventionally considered to be high. Power for any given test depends on the difference between μμ between groups/treatments, αα, nn, and σσ.

##Effect Size
Generally speaking, an effect size is a quantitative measure of the strength of a phenomenon. Here, we are interested in comparing two sample means, and the most common way to describe the effect size is as a standardized difference between the means of the groups being compared. In this case, we divide the difference between the means by the standard deviation: |(μ0|(μ0 - μA)|/σμA)|/σ. This results in a scaleless measure. Conventionally, effect sizes of 0.2 or less are considered to be low and of 0.8 or greater are considered to be high.

###Graphical Depiction of Power and Effect Size
```{r}
library(ggplot2)
library(manipulate)
power.plot <- function(sigma, muA, mu0, n, alpha, alternative = "two.tailed") {
    pow <- 0
    z <- (muA - mu0)/(sigma/sqrt(n))
    g <- ggplot(data.frame(mu = c(min(mu0 - 4 * sigma/sqrt(n), muA - 4 * sigma/sqrt(n)), 
        max(mu0 + 4 * sigma/sqrt(n), muA + 4 * sigma/sqrt(n)))), aes(x = mu)) + 
        ggtitle("Explore Power for Z Test")
    g <- g + ylim(c(0, max(dnorm(mu0, mu0, sigma/sqrt(n)) + 0.1, dnorm(muA, 
        muA, sigma/sqrt(n)) + 0.1)))
    g <- g + stat_function(fun = dnorm, geom = "line", args = list(mean = mu0, 
        sd = sigma/sqrt(n)), size = 1, col = "red", show.legend = TRUE)
    g <- g + stat_function(fun = dnorm, geom = "line", args = list(mean = muA, 
        sd = sigma/sqrt(n)), size = 1, col = "blue", show.legend = TRUE)
    
    if (alternative == "greater") {
        if (z > 0) {
            xcrit = mu0 + qnorm(1 - alpha) * sigma/sqrt(n)
            g <- g + geom_segment(x = xcrit, y = 0, xend = xcrit, yend = max(dnorm(mu0, 
                mu0, sigma/sqrt(n)) + 0.025, dnorm(muA, muA, sigma/sqrt(n)) + 
                0.025), size = 0.5, linetype = 3)
            g <- g + geom_polygon(data = data.frame(cbind(x = c(xcrit, seq(from = xcrit, 
                to = muA + 4 * sigma/sqrt(n), length.out = 100), muA + 4 * sigma/sqrt(n)), 
                y = c(0, dnorm(seq(from = xcrit, to = muA + 4 * sigma/sqrt(n), 
                  length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0))), 
                aes(x = x, y = y), fill = "blue", alpha = 0.5)
            pow <- pnorm(muA + 4 * sigma/sqrt(n), muA, sigma/sqrt(n)) - pnorm(xcrit, 
                muA, sigma/sqrt(n))
        }
    }
    if (alternative == "less") {
        if (z < 0) {
            xcrit = mu0 - qnorm(1 - alpha) * sigma/sqrt(n)
            g <- g + geom_segment(x = xcrit, y = 0, xend = xcrit, yend = max(dnorm(mu0, 
                mu0, sigma/sqrt(n)) + 0.025, dnorm(muA, muA, sigma/sqrt(n)) + 
                0.025), size = 0.5, linetype = 3)
            g <- g + geom_polygon(data = data.frame(cbind(x = c(muA - 4 * sigma/sqrt(n), 
                seq(from = muA - 4 * sigma/sqrt(n), to = xcrit, length.out = 100), 
                xcrit), y = c(0, dnorm(seq(from = muA - 4 * sigma/sqrt(n), to = xcrit, 
                length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0))), aes(x = x, 
                y = y), fill = "blue", alpha = 0.5)
            pow <- pnorm(xcrit, muA, sigma/sqrt(n)) - pnorm(muA - 4 * sigma/sqrt(n), 
                muA, sigma/sqrt(n))
        }
    }
    if (alternative == "two.tailed") {
        if (z > 0) {
            xcrit = mu0 + qnorm(1 - alpha/2) * sigma/sqrt(n)
            g <- g + geom_segment(x = xcrit, y = 0, xend = xcrit, yend = max(dnorm(mu0, 
                mu0, sigma/sqrt(n)) + 0.025, dnorm(muA, muA, sigma/sqrt(n)) + 
                0.025), size = 0.5, linetype = 3)
            g <- g + geom_polygon(data = data.frame(cbind(x = c(xcrit, seq(from = xcrit, 
                to = muA + 4 * sigma/sqrt(n), length.out = 100), muA + 4 * sigma/sqrt(n)), 
                y = c(0, dnorm(seq(from = xcrit, to = muA + 4 * sigma/sqrt(n), 
                  length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0))), 
                aes(x = x, y = y), fill = "blue", alpha = 0.5)
            pow <- pnorm(muA + 4 * sigma/sqrt(n), muA, sigma/sqrt(n)) - pnorm(xcrit, 
                muA, sigma/sqrt(n))
        }
        if (z < 0) {
            xcrit = mu0 - qnorm(1 - alpha/2) * sigma/sqrt(n)
            g <- g + geom_segment(x = xcrit, y = 0, xend = xcrit, yend = max(dnorm(mu0, 
                mu0, sigma/sqrt(n)) + 0.025, dnorm(muA, muA, sigma/sqrt(n)) + 
                0.025), size = 0.5, linetype = 3)
            g <- g + geom_polygon(data = data.frame(cbind(x = c(muA - 4 * sigma/sqrt(n), 
                seq(from = muA - 4 * sigma/sqrt(n), to = xcrit, length.out = 100), 
                xcrit), y = c(0, dnorm(seq(from = muA - 4 * sigma/sqrt(n), to = xcrit, 
                length.out = 100), mean = muA, sd = sigma/sqrt(n)), 0))), aes(x = x, 
                y = y), fill = "blue", alpha = 0.5)
            pow <- pnorm(xcrit, muA, sigma/sqrt(n)) - pnorm(muA - 4 * sigma/sqrt(n), 
                muA, sigma/sqrt(n))
        }
    }
    g <- g + annotate("text", x = max(mu0, muA) + 2 * sigma/sqrt(n), y = max(dnorm(mu0, 
        mu0, sigma/sqrt(n)) + 0.075, dnorm(muA, muA, sigma/sqrt(n)) + 0.075), 
        label = paste("Effect Size = ", round((muA - mu0)/sigma, digits = 3), 
            "\nPower = ", round(pow, digits = 3), sep = ""))
    g <- g + annotate("text", x = min(mu0, muA) - 2 * sigma/sqrt(n), y = max(dnorm(mu0, 
        mu0, sigma/sqrt(n)) + 0.075, dnorm(muA, muA, sigma/sqrt(n)) + 0.075), 
        label = "Red = mu0\nBlue = muA")
    g
}
```

```{r}
manipulate(power.plot(sigma, muA, mu0, n, alpha, alternative), sigma = slider(1, 
    10, step = 1, initial = 4), muA = slider(-10, 10, step = 1, initial = 2), 
    mu0 = slider(-10, 10, step = 1, initial = 0), n = slider(1, 50, step = 1, 
        initial = 16), alpha = slider(0.01, 0.1, step = 0.01, initial = 0.05), 
    alternative = picker("two.tailed", "greater", "less"))
```

```{r}
library(ggplot2)
library(manipulate)
power.test <- function(mu0, muA, sigma, alpha = 0.05, type, alternative) {
    p <- 0
    for (i in 2:200) {
        x <- power.t.test(n = i, delta = abs(muA - mu0), sd = sigma, sig.level = alpha, 
            power = NULL, type = type, alternative = alternative)
        p <- c(p, x$power)
    }
    d <- data.frame(cbind(1:200, p, 1 - p))
    critn <- 0
    for (i in 1:199) {
        if (p[i] < 0.8 && p[i + 1] >= 0.8) {
            critn <- i + 1
        } else {
            critn <- critn
        }
    }
    names(d) <- c("n", "power", "beta")
    g <- ggplot(data = d) + xlab("sample size n") + ylab("Type II Error Rate, Beta  (Red)\nand\nPower, 1-Beta (Blue)") + 
        ggtitle("Power for T Tests\n(assuming equal n and variance across the two groups)") + 
        ylim(0, 1) + geom_point(aes(x = n, y = power), colour = "blue", alpha = 1/2) + 
        geom_line(aes(x = n, y = power), colour = "blue", alpha = 1/2) + geom_line(aes(x = n, 
        y = 0.8), colour = "red", lty = 3) + geom_point(aes(x = n, y = beta), 
        colour = "red", alpha = 1/2) + geom_line(aes(x = n, y = beta), colour = "red", 
        alpha = 1/2) + geom_linerange(aes(x = critn, ymin = 0, ymax = 0.8), 
        colour = "blue", alpha = 1/4) + annotate("text", x = 150, y = 0.5, label = paste("Effect Size = ", 
        round(abs(mu0 - muA)/sigma, digits = 3), "\nCritical n = ", critn, sep = ""))
    print(g)
}
```

```{r}
manipulate(power.test(mu0, muA, sigma, alpha, type, alternative), mu0 = slider(-10, 
    10, initial = 3, step = 1), muA = slider(-10, 10, initial = 0, step = 1), 
    sigma = slider(1, 10, initial = 3, step = 1), alpha = slider(0.01, 0.1, 
        initial = 0.05, step = 0.01), alternative = picker("two.sided", "one.sided"), 
    type = picker("two.sample", "one.sample", "paired"))
```

